Conversations in time: interactive visualisation to explore structured temporal data

### 2020-09-25 tsibbletalk.tex
minor corrections

\abstract{%
Temporal data often has a hierarchical structure, defined by categorical
variables describing different levels, such as political regions or
sales products. Nesting of categorical variables produces a hierarchical
structure. The \CRANpkg{tsibbletalk} package is developed to allow a
user to interactively explore temporal data, relative to the nested or
crossed structures. It can help to discover differences between category
levels, and uncover interesting periodic or aperiodic slices. The
package implements a shared \texttt{tsibble} object that allows for
linked brushing between coordinated views, and a shiny module that aids
in wrapping time lines for seasonal patterns. The tools are demonstrated
using two data examples: domestic tourism in Australia and foot traffic
in Melbourne.
}

@@

Temporal data typically arrives as a set of many observational units
measured over time. Some variables may be categorical, indicating a
hierarchy in the collection process, that may be measurements taken in
different geographic regions, or types of products sold by the one
company. Exploring these multiple features can be daunting. Ensemble
graphics \citep{unwin_ensemble_2018} bundle multiple views of a data set
together into one composite figure. These provide an effective approach
for exploring and digesting many different aspects of temporal data.
Adding interactivity to the ensemble can greatly enhance the exploration
process.

This article describes new software, the \CRANpkg{tsibbletalk} package,
for exploring temporal data using linked views and time wrapping. The
next section provides some background to the approach based on setting
up data structures and workflow. The section following explains the
mechanism for constructing interactivity, to link between multiple
hierarchical data objects and hence plots. The section after this
describes the set up for interactively slicing and dicing time to wrap a
series on itself to investigate periodicities.

@@

The \CRANpkg{tsibble} package \citep{wang2020tsibble} introduced a
unified temporal data structure, referred to as a \texttt{tsibble}, to
represent time series and longitudinal data in a tidy format
\citep{wickham2014tidy}. A \texttt{tsibble} extends the
\texttt{data.frame} and \CRANpkg{tibble} classes with the temporal
contextual metadata: \texttt{index} and \texttt{key}. The \texttt{index}
declares a data column that holds time-related indices. The \texttt{key}
identifies a collection of related series or panels observed over the
\texttt{index}-defined period, which can comprise multiple columns. An
example of a \texttt{tsibble} can be found in the monthly Australian
retail trade turnover data (\texttt{aus\_retail}), available in the
\CRANpkg{tsibbledata} package \citep{R-tsibbledata}, shown below. The
\texttt{Month} column holds year-months as the \texttt{index}.
\texttt{State} and \texttt{Industry}, together, are the identifiers for
these 152 series, which form the \texttt{key}. Note that the column
\texttt{Series\ ID} could be an alternative option for setting up the
\texttt{key}, but \texttt{State} and \texttt{Industry} are more readable
and informative. The \texttt{index} and \texttt{key} are ``sticky''
columns to a \texttt{tsibble}, forming critical pieces for fluent
downstream temporal data analysis.

@@

Figure \ref{fig:highlight-retail} illustrates examining temporal data
with many keys. The data has 152 series corresponding to different items
in retail data. The multiple series are displayed using an overlaid time
series plot, along with a scatterplot of two variables (trend versus
seasonal strength) from feature space, where each series is represented
by a dot. The feature space is computed using the \texttt{features()}
function from \CRANpkg{feasts}, which summarises the original data for
each series using various statistical features. This function along with
other \textbf{tidyverts} functions is \texttt{tsibble}-aware, and
outputs a table in a reduced form where each row corresponds to a
series, which can be graphically displayed as in Figure
\ref{fig:highlight-retail-2}.

@@

R \CRANpkg{shiny} \citep{R-shiny} and \CRANpkg{htmlwidgets}
\citep{R-htmlwidgets} provides and infrastructure connecting R with HTML
elements and JavaScript that provide the interactivity. The
\CRANpkg{htmlwidgets} package makes it possible to embed JavaScript
libraries into R so that users are able to write only R code to generate
web-based plots. Many JavaScript charting libraries have been ported to
R as HTML widgets, including \CRANpkg{plotly} \citep{plotly2020},
\CRANpkg{rbokeh} \citep{R-rbokeh}, and \CRANpkg{leaflet}
\citep{R-leaflet} for maps. Interactions between different widgets can
be achieved with \CRANpkg{shiny} or \CRANpkg{crosstalk}
\citep{R-crosstalk}. The \CRANpkg{crosstalk} extends
\CRANpkg{htmlwidgets} with shared R6 instances to support linked
brushing and filtering across widgets, without relying on
\CRANpkg{shiny}.

@@

The \CRANpkg{tsibbletalk} package introduces a shared tsibble instance
built on a \texttt{tsibble}. This allows for seamless communication
between different plots of temporal data. The
\texttt{as\_shared\_tsibble()} function turns a \texttt{tsibble} into a
shared instance, \texttt{SharedTsibbleData}, which is a subclass of
\texttt{SharedData} from \CRANpkg{crosstalk}). This is the object
driving data transmission across multiple views. The
\CRANpkg{tsibbletalk} package aims to streamline interactive exploration
of temporal data, with the focus of temporal elements and structured
linking.

@@

As opposed to one-to-one linking, \CRANpkg{tsibbletalk} defaults to
categorical variable linking where selecting one or more observations in
one category will broadcast to all other observations in this category.
In addition, linking is also by time: within the time series plot, click
on any data point, and whole line will be highlighted in response. The
\texttt{as\_shared\_tsibble()} uses \texttt{tsibble}'s \texttt{key}
variables to achieve these types of linking.

The approach can also accommodate hierarchical temporal data. When one
key variable is nested within another, such as regional areas within a
state, this is considered to be a hierarchical structure. These time
series are referred to as hierarchical and grouped time series in the
literature \citep{fpp}. The \texttt{aus\_retail} above is an example.
Each series in the data corresponds to all possible combinations of the
\texttt{State} and \texttt{Industry} variables, which means they are
intrinsically crossed with each other.

The \texttt{spec} argument in \texttt{as\_shared\_tsibble()}
constructing hybrid linking, that incorporates hierarchical and
categorical linking. A symbolic formula can be passed to the
\texttt{spec} argument, to incorporate structured specifications in the
\texttt{key}. Adopting \citet{Wilkinson1973}'s notation for factorial
models, the \texttt{spec} follows the \texttt{/} and \texttt{*} operator
conventions to declare nesting and crossing variables, respectively. The
\texttt{spec} for the \texttt{aus\_retail} data is therefore specified
as \texttt{State\ *\ Industry} or \texttt{Industry\ *\ State}, which is
the default for the presence of multiple \texttt{key} variables. If
there is a hierarchy in the data, using \texttt{/} is required to
indicate the parent-child relation, for a strictly one directional
\texttt{parent/child}.

@@

In summary, shared the tsibble data nicely bridges between the
\CRANpkg{crosstalk} and \textbf{tidyverts} ecosystems for temporal data
using the common ``key''. The \texttt{as\_shared\_tsibble()} provides a
symbolic user interface for effortless construction of a hybrid of
hierarchical and categorical linking between plots. The
\texttt{plotly\_key\_tree()} function, in turn, decodes the hierarchical
specification to plot a tree for data overview and navigation, when
accompanied by more detailed plots.

@@

Figure \ref{fig:wrap-ped-2} and \ref{fig:wrap-ped-3} show slicing and
wrapping of the series into daily and weekly sections, respectively.
Multiple seasonality pops out. There tends to be a daily pattern,
especially visible at the main train station, Southern Cross. There is
also a week day vs weekend pattern, also most visible at Southern Cross.
These seasonal patterns are still present during the lockdown, but the
magnitude is greatly reduced. Numbers are also down at the produce
market and the shopping center. Birrarung Marr is the most affected.
This is the location of special events and its clear that these have
completely disappeared during lockdown.

The wrapping procedure involves slicing the time index into seasonal
periods of interest, and the result is diced time. For example, hourly
pedestrian data can be decomposed into 24-hour blocks, which then
overlays the counts for all respective days, as done in plot
\ref{fig:wrap-ped-2}. For exploration, this slice position should be
controlled interactively, so that many different slices can be examined
rapidly. This can be achieved using shiny, with the functions provided
in a \CRANpkg{tsibbletalk}.

This shiny module, decoupled to \texttt{tsibbleWrapUI()} and
\texttt{tsibbleWrapServer()}, presents a clean interface and forms a
reusable code that could be embedded in any shiny application. In
general, a shiny module provides a vehicle for modularising shiny
applications, relevant for both users and developers. As with all shiny
modules, the first argument in both functions in \CRANpkg{tsibbletalk}
requires a user-supplied session id that must be unique. The UI function
\texttt{tsibbleWrapUI()} simply shows a slider that animates or controls
the number of periods to be diced. The workhorse is the server function
\texttt{tsibbleWrapServer()}, encapsulating the algorithm that
transforms data and sends messages to update the plot accordingly. The
\texttt{plot} argument expects a \texttt{ggplot} or \texttt{plotly}
object, where one can plot data using either lines or other graphical
elements (such as boxplots). As the function name suggests, a (shared)
tsibble is needed to start the engine, so that the time \texttt{index}
can be retrieved for dissection. The \texttt{period} option semantically
takes a desired number of seasonal periods to be shifted, for example
data shifted by ``1 day'', ``2 days'', or ``1 week'', etc. In other
words, the \texttt{period} defines the grind level. For date-times
(represented by \texttt{POSIXt}), the granularity ranges from fine
``day'' to a much coarser ``year''. The following code snippet generates
Figure \ref{fig:wrap-ped}.

>>>>

\abstract{%
Temporal data often has a hierarchical structure, defined by categorical
variables describing different levels, such as political regions or
sales products. Nesting of categorical variables produces a hierarchical
structure. The \CRANpkg{tsibbletalk} package is developed to allow a
user to interactively explore temporal data, relative to the nested or
crossed structures. It can help to discover differences between category
levels, and uncover interesting periodic or aperiodic slices. The
package implements a shared \texttt{tsibble} object that allows for
linked brushing between coordinated views, and a shiny module that aids
in wrapping time lines for seasonal patterns. The tools are demonstrated
using two data examples: domestic tourism in Australia and pedestrian
traffic in Melbourne.
}

@@

Temporal data typically arrives as a set of many observational units
measured over time. Some variables may be categorical, containing a
hierarchy in the collection process, that may be measurements taken in
different geographic regions, or types of products sold by the one
company. Exploring these multiple features can be daunting. Ensemble
graphics \citep{unwin_ensemble_2018} bundle multiple views of a data set
together into one composite figure. These provide an effective approach
for exploring and digesting many different aspects of temporal data.
Adding interactivity to the ensemble can greatly enhance the exploration
process.

This paper describes new software, the \CRANpkg{tsibbletalk} package,
for exploring temporal data using linked views and time wrapping. We
first provide some background to the approach based on setting up data
structures and workflow, and give an overview of interactive systems in
R. The section following introduces the \CRANpkg{tsibbletalk} package.
We explain the mechanism for constructing interactivity, to link between
multiple hierarchical data objects and hence plots, and describe the set
up for interactively slicing and dicing time to wrap a series on itself
to investigate periodicities.

@@

The \CRANpkg{tsibble} package \citep{wang2020tsibble} introduced a
unified temporal data structure, referred to as a \texttt{tsibble}, to
represent time series and longitudinal data in a tidy format
\citep{wickham2014tidy}. A \texttt{tsibble} extends the
\texttt{data.frame} and \texttt{tibble} classes with the temporal
contextual metadata: \texttt{index} and \texttt{key}. The \texttt{index}
declares a data column that holds time-related indices. The \texttt{key}
identifies a collection of related series or panels observed over the
\texttt{index}-defined period, which can comprise multiple columns. An
example of a \texttt{tsibble} can be found in the monthly Australian
retail trade turnover data (\texttt{aus\_retail}), available in the
\CRANpkg{tsibbledata} package \citep{R-tsibbledata}, shown below. The
\texttt{Month} column holds year-months as the \texttt{index}.
\texttt{State} and \texttt{Industry} are the identifiers for these 152
series, which form the \texttt{key}. Note that the column
\texttt{Series\ ID} could be an alternative option for setting up the
\texttt{key}, but \texttt{State} and \texttt{Industry} are more readable
and informative. The \texttt{index} and \texttt{key} are ``sticky''
columns to a \texttt{tsibble}, forming critical pieces for fluent
downstream temporal data analysis.

@@

Figure \ref{fig:highlight-retail} illustrates examining temporal data
with many keys. The data has 152 series corresponding to different
industries in retail data. The multiple series are displayed using an
overlaid time series plot, along with a scatterplot of two variables
(trend versus seasonal strength) from feature space, where each series
is represented by a dot. The feature space is computed using the
\texttt{features()} function from \CRANpkg{feasts}, which summarises the
original data for each series using various statistical features. This
function along with other \textbf{tidyverts} functions is
\texttt{tsibble}-aware, and outputs a table in a reduced form where each
row corresponds to a series, which can be graphically displayed as in
Figure \ref{fig:highlight-retail-2}.

@@

R \CRANpkg{shiny} \citep{R-shiny} and \CRANpkg{htmlwidgets}
\citep{R-htmlwidgets} provide infrastructure connecting R with HTML
elements and JavaScript that support the interactivity. The
\CRANpkg{htmlwidgets} package makes it possible to embed JavaScript
libraries into R so that users are able to write only R code to generate
web-based plots. Many JavaScript charting libraries have been ported to
R as HTML widgets, including \CRANpkg{plotly} \citep{plotly2020},
\CRANpkg{rbokeh} \citep{R-rbokeh}, and \CRANpkg{leaflet}
\citep{R-leaflet} for maps. Interactions between different widgets can
be achieved with \CRANpkg{shiny} or \CRANpkg{crosstalk}
\citep{R-crosstalk}. The \CRANpkg{crosstalk} extends
\CRANpkg{htmlwidgets} with shared R6 instances to support linked
brushing and filtering across widgets, without relying on
\CRANpkg{shiny}.

@@

The \CRANpkg{tsibbletalk} package introduces a shared tsibble instance
built on a \texttt{tsibble}. This allows for seamless communication
between different plots of temporal data. The
\texttt{as\_shared\_tsibble()} function turns a \texttt{tsibble} into a
shared instance, \texttt{SharedTsibbleData}, which is a subclass of
\texttt{SharedData} from \CRANpkg{crosstalk}. This is the object driving
data transmission across multiple views. The \CRANpkg{tsibbletalk}
package aims to streamline interactive exploration of temporal data,
with the focus of temporal elements and structured linking.

@@

As opposed to one-to-one linking, \CRANpkg{tsibbletalk} defaults to
categorical variable linking where selecting one or more observations in
one category will broadcast to all other observations in this category.
That is, linking is by time: within the time series plot, click on any
data point, and whole line will be highlighted in response. The
\texttt{as\_shared\_tsibble()} uses \texttt{tsibble}'s \texttt{key}
variables to achieve these types of linking.

The approach can also accommodate temporal data of nesting and crossing
structures. These time series are referred to as hierarchical and
grouped time series in the literature \citep{fpp}. The
\texttt{aus\_retail} above is an example of grouped time series. Each
series in the data corresponds to all possible combinations of the
\texttt{State} and \texttt{Industry} variables, which means they are
intrinsically crossed with each other. When one key variable is nested
within another, such as regional areas within a state, this is
considered to be a hierarchical structure.

The \texttt{spec} argument in \texttt{as\_shared\_tsibble()} provides a
means to construct hybrid linking, that incorporates hierarchical and
categorical linking. A symbolic formula can be passed to the
\texttt{spec} argument, to allow for structured specifications in the
\texttt{key}. Adopting \citet{Wilkinson1973}'s notation for factorial
models, the \texttt{spec} follows the \texttt{/} and \texttt{*} operator
conventions to declare nesting and crossing variables, respectively. The
\texttt{spec} for the \texttt{aus\_retail} data is therefore specified
as \texttt{State\ *\ Industry} or \texttt{Industry\ *\ State}, which is
the default for the presence of multiple \texttt{key} variables. If
there is a hierarchy in the data, using \texttt{/} is required to
indicate the parent-child relation, for a strictly one directional
\texttt{parent/child}.

@@

In summary, shared tsibble data nicely bridges between the
\CRANpkg{crosstalk} and \textbf{tidyverts} ecosystems for temporal data
using the common ``key''. The \texttt{as\_shared\_tsibble()} provides a
symbolic user interface for effortless construction of a hybrid of
hierarchical and categorical linking between plots. The
\texttt{plotly\_key\_tree()} function, in turn, decodes the hierarchical
specification to plot a tree for data overview and navigation, when
accompanied by more detailed plots.

@@

Figure \ref{fig:wrap-ped-2} and \ref{fig:wrap-ped-3} show slicing and
wrapping of the series into daily and weekly sections, respectively.
Multiple seasonalities pop out. There tends to be a daily pattern,
especially visible at the main train station, Southern Cross Station.
There is also a weekday vs weekend pattern, also most visible at
Southern Cross Station. These seasonal patterns are still present during
the lockdown, but the magnitude is greatly reduced. Numbers are also
down at the produce market and the shopping center. Birrarung Marr is
the most affected. This is the location of special events, and it is
clear that these have completely disappeared during lockdown.

The wrapping procedure involves slicing the time index into seasonal
periods of interest, and the result is diced time. For example, hourly
pedestrian data can be decomposed into 24-hour blocks, which then
overlays the counts for all respective days, as done in plot
\ref{fig:wrap-ped-2}. For exploration, this slice position should be
controlled interactively, so that many different slices can be examined
rapidly. This can be achieved using shiny, with the functions provided
in the \CRANpkg{tsibbletalk}.

This shiny module, decoupled to \texttt{tsibbleWrapUI()} and
\texttt{tsibbleWrapServer()}, presents a clean interface and forms a
reusable component that could be embedded in any shiny application. In
general, a shiny module provides a vehicle for modularising shiny
applications, relevant for both users and developers. As with all shiny
modules, the first argument in both functions in \CRANpkg{tsibbletalk}
requires a user-supplied session id that must be unique. The UI function
\texttt{tsibbleWrapUI()} simply shows a slider that animates or controls
the number of periods to be diced. The workhorse is the server function
\texttt{tsibbleWrapServer()}, encapsulating the algorithm that
transforms data and sends messages to update the plot accordingly. The
\texttt{plot} argument expects a \texttt{ggplot} or \texttt{plotly}
object, where one can plot data using either lines or other graphical
elements (such as boxplots). As the function name suggests, a (shared)
tsibble is needed to start the engine, so that the time \texttt{index}
can be retrieved for dissection. The \texttt{period} option semantically
takes a desired number of seasonal periods to be shifted, for example
data shifted by ``1 day'', ``2 days'', or ``1 week'', etc. In other
words, the \texttt{period} defines the grind level. For date-times
(represented by \texttt{POSIXt}), the granularity ranges from fine
``day'' to a much coarser ``year''. The following code snippet generates
Figure \ref{fig:wrap-ped}.
